A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-29 bioc-workflow.Rmd
fix wrong citation

If you have questions about this workflow in particular, or about
other Bioconductor packages in general, they are best asked on the
[Bioconductor support site](https://support.bioconductor.org/)
following the
[posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). Questions
can be tagged with specific package names or keywords. For more
general information about mass spectrometry and proteomics, the
readers are invited to read the `r Biocexptpkg("RforProteomics")`
package vignettes and associated papers [@Gatto:2014;@Gatto2015].

>>>>

If you have questions about this workflow in particular, or about
other Bioconductor packages in general, they are best asked on the
[Bioconductor support site](https://support.bioconductor.org/)
following the
[posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). Questions
can be tagged with specific package names or keywords. For more
general information about mass spectrometry and proteomics, the
readers are invited to read the `r Biocexptpkg("RforProteomics")`
package vignettes and associated papers [@Gatto:2014;@Gatto:2015].

### 2016-09-29 bioc-workflow.tex
fix wrong citation

In this workflow, we will describe how to prepare the spatial proteomics
data starting from a spreadsheet containing quantitative mass
spectrometry data. We will focus on a recent pluripotent mouse embryonic
stem cells experiment {[}@hyper{]}. Additional annotated and
pre-formatted datasets from various species are readily available in the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
package.

@@

If you have questions about this workflow in particular, or about other
Bioconductor packages in general, they are best asked on the
\href{https://support.bioconductor.org/}{Bioconductor support site}
following the
\href{http://www.bioconductor.org/help/support/posting-guide/}{posting
guidelines}. Questions can be tagged with specific package names or
keywords. For more general information about mass spectrometry and
proteomics, the readers are invited to read the
\emph{\href{http://bioconductor.org/packages/RforProteomics}{RforProteomics}}
package vignettes and associated papers {[}@Gatto:2014;@Gatto2015{]}.

@@

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
{[}@hyper{]}. The data was generated using hyperplexed LOPIT
(hyperLOPIT), a state-of-the-art method relying on improved sub-cellular
fractionation and more accurate quantitation, leading to more reliable
classification of protein localisation across the whole sub-cellular
space. The method uses an elaborate sub-cellular fractionation scheme,
enabled by the use of Tandem Mass Tag (TMT) {[}Thompson:2003{]} 10-plex
and application of the MS data acquisition technique named synchronous
precursor selection MS\textsuperscript{3} (SPS-MS\textsuperscript{3})
{[}@McAlister:2014{]}, for TMT quantification with high accuracy and
precision. Three biological replicates were generated from the E14TG2a
experiment, the first was to target low density fractions and the second
and third were to emphasis separation of the denser organelles. The
intersect of replicates 1 and 2 was treated as a 20-plex dataset for the
analysis. As discussed in the manuscript {[}@hyper{]} as it has been
shown that combining replicates across from different gradients can
increase spatial resolution {[}@Trotter:2010{]}. The combination of
replicates resulted in 5032 proteins common in both experiments.

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2} in
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}. The
function takes a single spreadsheet as input and extracts the columns
containing the quantitation data, as identified by the argument
\texttt{ecol}, to create the expression data, while the other columns in
the spreadsheet are appended to the feature meta-data slot. By example,
in the code chunk below we read in the \texttt{csv} spreadsheet
containing the quantitation data from the intersect of replicates 1 and
2 of the mouse map {[}@hyper{]}, using the \texttt{readMSnSet2}
function. The data is as available online with the manuscript (see tab 2
of the \texttt{xlsx} supplementary data set 1 in {[}@hyper{]}, which
should be exported as a text-based spreadsheet). It is also available as
a \texttt{csv} in the Bioconductor \texttt{r Biocexptpkg("pRolocdata")}
data package.

@@

Throughout this workflow we refer to the different columns that are
found in the \texttt{exprs} (expression data) slot as channels (short
for TMT channels). In the frame of LOPIT and hyperLOPIT these channels
consititue the relative abundance of each protein (along the rows) in
the channel of interest. Each TMT channel originates from fractions
collected from the density gradient, or a set of pooled fractions or may
be a sample originating from an alternative preparation e.g.~such as
from the chromatin enrichment performed in Christoforou et al
{[}@hyper{]}. Information about which gradient fractions were used for
which tag should also be stored in the sample meta-data \texttt{pData}
slot.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Thu Sep 29 10:28:15 2016 
##  MSnbase version: 1.21.7
\end{verbatim}

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow
{[}@Webb-Robertson:2015;@Lazar:2016{]}. In {[}@Gatto:2014b{]}, we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. Missing data can be imputated using
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}'s
\texttt{impute} function. The method underlying the imputation method is
then determined by a \texttt{methods} parameter. In our particular case,
missing values are indicative of protein groups that were not profiles
in both replicates.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Thu Sep 29 10:28:15 2016 
## Subset [6725,20][5032,20] Thu Sep 29 10:28:15 2016 
## Removed features with more than 0 NAs: Thu Sep 29 10:28:15 2016 
## Dropped featureData's levels Thu Sep 29 10:28:15 2016 
##  MSnbase version: 1.21.7
\end{verbatim}

@@

Data quality is routinely examined through visualisation to verify that
sub-cellular niches have been separated along the gradient. Based on De
Duve's principle {[}@DeDuve:1981{]} proteins that co-localise exhibit
similar quantitation profiles across the gradient fractions employed.
One approach that has been widely used to visualise and inspect high
throughput mass spectrometry-based proteomics data is principal
components analysis (PCA). PCA is one of many dimensionality reduction
methods, that allow one to effectively summarise multi-dimensional data
in to 2 or 3 dimensions to enable visualisation. Very generally, the
original continuous multi-dimensional data is transformed into a set of
orthogonal components ordered according to the amount of variability
that they describe. The \texttt{plot2D} method in
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} allows one
to plot the principal components (PCs) of a dataset against one another,
by default the first two components are plotted on the x- and y-axis,
respectively (the \texttt{dims} argument can be used to plot other PCs).
If distinct clusters are observed, we assume that there is organellar
separation present in the data. Although, representing the
multi-dimensional data along a limited set of PCs does not give us a
hard quantitative measure of separation, it is extremely useful
summarising complex experimental information in one figure, to get an
simplified overview of the data.

@@

In the context of spatial proteomics, a marker protein is defined as a
well-known resident of a specific sub-cellular niche in a species
\emph{and} condition of interest. Applying this to machine learning
(ML), and specifically supervised learning, for the task of protein
localisation prediction, these markers constitute the labelled training
data to use as input to a classification analyses. Defining well-known
residents, and obtaining labelled training data for ML analyses can be
time consuming, but it is important to define markers that are
representative of the multivariate data space and on which a classifier
will be trained and generated.
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} provides a
convenience function, \texttt{addMarkers}, to directly add markers to a
\texttt{MSnSet} object, as demonstrated in the code chunk below. These
marker sets can be accessed using the \texttt{pRolocmarkers()} function.
Marker sets are stored as a simple named vector in R, and originate from
in-house user-defined sets of markers or from previous published studies
{[}@Gatto:2014b{]}, which are continuosly updated and integrated.

@@

As demonstrated in {[}@hyper{]} and illustrated in the PCA plot above,
the Golgi apparatus proteins display a dynamic pattern, noting sets of
Golgi marker proteins that are distributed amongst other subcellular
structures, an observation supported by microscopy. As such, we are
going to reset the annotation of Golgi markers to unknown using the
\texttt{fDataTounknown} function. It is often used to replace empty
strings (``'') or missing values in the markers definition to a common
definition of \emph{unknown localisation}.

@@

In general, the Gene Ontology (GO) {[}@Ashburner:2000{]}, and in
particular the cellular compartment (CC) namespace are a good starting
point for protein annotation and marker definition. It is important to
note however that automatic retrieval of sub-cellular localisation
information, from
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} or
elsewhere, is only the beginning in defining a marker set for downstream
analyses. Expert curation is vital to check that any annotation added is
in the correct context for the the biological question under
investigation.

@@

With the aim of maximising the sub-cellular resolution and,
consequently, the reliability in protein sub-cellular assignments, we
follow the advice in {[}@Trotter:2010{]} and combine replicated spatial
proteomics experiments as described above. Indeed, Trotter et al. have
shown a significant improvement in protein--organelle association upon
direct combination of single experiments, in particular when these
resolve different subcellular niches.

@@

The extraction of sub-cellular protein clusters can be difficult owing
to the limited number of marker proteins that exist in databases and
elsewhere. Furthermore, given the vast complexity of the cell, automatic
annotation retrieval does not always give a full representation of the
true sub-cellular diversity in the data. For downstream analyses, such
as supervised machine learning, it is desirable to obtain reliable
markers that cover as many sub-cellular niches as possible, as these
markers are directly used in the training phase of the ML
classification. We find that a lack of sub-cellular diversity in the
labelled training data leads to prediction errors, as unlabelled
instances can only be assigned to a class that exists in the training
data {[}@Breckels:2013{]}. In such scenarios novelty detection can be
useful to identify data-specific sub-cellular groupings such as
organelles and protein complexes. The phenotype discovery (phenoDisco)
algorithm {[}@Breckels:2013{]} is one such method and is available in
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}}. It is an
iterative semi-supervised learning method that combines the
classification of proteins on existing labelled data with the detection
of new clusters.

In addition to extracting new phenotypes, novelty detection methods are
also useful for confirming the presence of known or postulated clusters
in an unbiased fashion. For example, in {[}@hyper{]} the
\texttt{phenoDisco} algorithm was used to confirm the data-specific
presence of the nucleus and nucleus sub-compartments. In the code chunk
below, we demonstrate how to do this analysis, highlighting some of the
optional arguments and parameters available for phenotype extraction and
give some advice on how to interpret the output.

As the \texttt{phenoDisco} algorithm is semi-supervised it uses both
labelled (markers) and unlabelled data to explore the data structure and
find new sub-cellular data clusters. Thus the first step is to define
some input labelled data i.e.~markers, that the algorithm will use as
input for the supervised learning aspect of the algorithm. As described
in {[}@hyper{]} we define a set of markers to use as input for the
analyses that cover well-known residents from three distinct organelle
structures; the mitochondria, plasma membrane and ER, and from three
well-known and abundant protein complexes; the proteasome and two
ribosomal subunits, 40S and 60S. These input markers are stored in the
\texttt{featureData} column of \texttt{hl} where
\texttt{fcol = "phenoDisco.Input"}. We can use the convenience accessor
function \texttt{getMarkers} to print out a table of the markers
contained in this marker set. These initial markers were manually
curated using information from the UniProt database, the Gene Ontology
and the literature.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Thu Sep 29 10:28:15 2016 
## Added markers from  'pdres' marker vector. Thu Sep 29 10:28:17 2016 
##  MSnbase version: 1.99.2
\end{verbatim}

@@

Supervised machine learning, also known as classification, is an
essential tool for the assignment of proteins to distinct sub-cellular
niches. Using a set of labelled training examples i.e.~markers, we can
train a machine learning classifier to learn a mapping between the data
i.e.~the quantitative protein profiles, and a known localisation. The
trained classifier can then be used to predict the localisation of a
protein of unknown localisation, based on its observed protein profile.
To date, this method has been extensively used in spatial quantitative
proteomics to assign thousands of proteins to distinct sub-cellular
niches {[}@hyper; @Groen:2014; @Trotter:2010; @Hall:2009; @Dunkley:2006;
@Tan:2009{]}.

@@

\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4356698               0.3028843               0.3825420 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6825826               0.5995243               0.4245311 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4034908               0.5882168               0.9501241 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7951459               0.7096694               0.3149301 
##         Plasma membrane              Proteasome 
##               0.7091667               0.4204304
\end{verbatim}

@@

\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      86                     169                      86 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     295                     479                      91 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      26                     125                     522 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     230                     344                      38 
##         Plasma membrane              Proteasome                 unknown 
##                     325                     159                    2057
\end{verbatim}

@@

In addition to high quality MS-based quantitative proteomics data there
exist a number of other sources of information that are freely available
in the public domain that may be useful to assign a protein to its
sub-cellular niche. For example, imaging from immunofluorescence
microscopy, protein annotations and sequences, and protein-protein
interactions among others, represent a rich and vast source of
complementary information. We can integrate this auxiliary information
with our primary MS-based quantitative data using a paradigm known as
transfer learning (TL). The integration of data between different
technologies is one of the biggest challenges in computational biology
to date and the
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} package
provides functionality to do such analyses. We recently developed two
transfer learning algorithms using a \emph{k}-NN and SVM framework and
applied them to the task of protein localisation prediction
{[}@Breckels:2016{]}. In this section we will begin with explaining the
concept of transfer learning and then show how to apply this in the
frame of spatial proteomics and protein localisation prediction.

@@

Now we have generated our auxiliary data we can use the \emph{k}-NN
implementation of transfer learning available in
\texttt{Biocpkg("pRoloc")} to integrate this with our primary MS-based
quantitative proteomics data using the functions
\texttt{knntlOptimisation} to estimate the free-parameters for the
integration, and \texttt{knntlClassification} to do the predictions. We
have shown that using transfer learning in the context of spatial
proteomics results in the assignment of proteins to sub-cellular niches
with a higher generalisation accuracy than using standard supervised
machine learning with a single source of information
{[}@Breckels:2016{]}.

>>>>

In this workflow, we will describe how to prepare the spatial proteomics
data starting from a spreadsheet containing quantitative mass
spectrometry data. We will focus on a recent pluripotent mouse embryonic
stem cells experiment (A. Christoforou et al. 2016). Additional
annotated and pre-formatted datasets from various species are readily
available in the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
package.

@@

If you have questions about this workflow in particular, or about other
Bioconductor packages in general, they are best asked on the
\href{https://support.bioconductor.org/}{Bioconductor support site}
following the
\href{http://www.bioconductor.org/help/support/posting-guide/}{posting
guidelines}. Questions can be tagged with specific package names or
keywords. For more general information about mass spectrometry and
proteomics, the readers are invited to read the
\emph{\href{http://bioconductor.org/packages/RforProteomics}{RforProteomics}}
package vignettes and associated papers (L. Gatto and Christoforou 2014;
Gatto et al. 2015).

@@

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a) (A.
Christoforou et al. 2016). The data was generated using hyperplexed
LOPIT (hyperLOPIT), a state-of-the-art method relying on improved
sub-cellular fractionation and more accurate quantitation, leading to
more reliable classification of protein localisation across the whole
sub-cellular space. The method uses an elaborate sub-cellular
fractionation scheme, enabled by the use of Tandem Mass Tag (TMT)
{[}Thompson:2003{]} 10-plex and application of the MS data acquisition
technique named synchronous precursor selection MS\textsuperscript{3}
(SPS-MS\textsuperscript{3}) (McAlister et al. 2014), for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was to
target low density fractions and the second and third were to emphasis
separation of the denser organelles. The intersect of replicates 1 and 2
was treated as a 20-plex dataset for the analysis. As discussed in the
manuscript (A. Christoforou et al. 2016) as it has been shown that
combining replicates across from different gradients can increase
spatial resolution (M. W. B. Trotter et al. 2010). The combination of
replicates resulted in 5032 proteins common in both experiments.

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2} in
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}. The
function takes a single spreadsheet as input and extracts the columns
containing the quantitation data, as identified by the argument
\texttt{ecol}, to create the expression data, while the other columns in
the spreadsheet are appended to the feature meta-data slot. By example,
in the code chunk below we read in the \texttt{csv} spreadsheet
containing the quantitation data from the intersect of replicates 1 and
2 of the mouse map (A. Christoforou et al. 2016), using the
\texttt{readMSnSet2} function. The data is as available online with the
manuscript (see tab 2 of the \texttt{xlsx} supplementary data set 1 in
(A. Christoforou et al. 2016), which should be exported as a text-based
spreadsheet). It is also available as a \texttt{csv} in the Bioconductor
\texttt{r Biocexptpkg("pRolocdata")} data package.

@@

Throughout this workflow we refer to the different columns that are
found in the \texttt{exprs} (expression data) slot as channels (short
for TMT channels). In the frame of LOPIT and hyperLOPIT these channels
consititue the relative abundance of each protein (along the rows) in
the channel of interest. Each TMT channel originates from fractions
collected from the density gradient, or a set of pooled fractions or may
be a sample originating from an alternative preparation e.g.~such as
from the chromatin enrichment performed in Christoforou et al (A.
Christoforou et al. 2016). Information about which gradient fractions
were used for which tag should also be stored in the sample meta-data
\texttt{pData} slot.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Thu Sep 29 13:42:11 2016 
##  MSnbase version: 1.21.7
\end{verbatim}

@@

Missing data are a recurrent issue in mass spectrometry applications,
and should be addressed independently of this workflow (Webb-Robertson
et al. 2015; Lazar et al. 2016). In (Gatto et al. 2014), we have
described how a high content in missing values in spatial proteomics
data and their inappropriate handling leads to a reduction of
sub-cellular resolution. Missing data can be imputated using
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}'s
\texttt{impute} function. The method underlying the imputation method is
then determined by a \texttt{methods} parameter. In our particular case,
missing values are indicative of protein groups that were not profiles
in both replicates.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Thu Sep 29 13:42:11 2016 
## Subset [6725,20][5032,20] Thu Sep 29 13:42:11 2016 
## Removed features with more than 0 NAs: Thu Sep 29 13:42:11 2016 
## Dropped featureData's levels Thu Sep 29 13:42:11 2016 
##  MSnbase version: 1.21.7
\end{verbatim}

@@

Data quality is routinely examined through visualisation to verify that
sub-cellular niches have been separated along the gradient. Based on De
Duve's principle (De Duve and Beaufay 1981) proteins that co-localise
exhibit similar quantitation profiles across the gradient fractions
employed. One approach that has been widely used to visualise and
inspect high throughput mass spectrometry-based proteomics data is
principal components analysis (PCA). PCA is one of many dimensionality
reduction methods, that allow one to effectively summarise
multi-dimensional data in to 2 or 3 dimensions to enable visualisation.
Very generally, the original continuous multi-dimensional data is
transformed into a set of orthogonal components ordered according to the
amount of variability that they describe. The \texttt{plot2D} method in
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} allows one
to plot the principal components (PCs) of a dataset against one another,
by default the first two components are plotted on the x- and y-axis,
respectively (the \texttt{dims} argument can be used to plot other PCs).
If distinct clusters are observed, we assume that there is organellar
separation present in the data. Although, representing the
multi-dimensional data along a limited set of PCs does not give us a
hard quantitative measure of separation, it is extremely useful
summarising complex experimental information in one figure, to get an
simplified overview of the data.

@@

In the context of spatial proteomics, a marker protein is defined as a
well-known resident of a specific sub-cellular niche in a species
\emph{and} condition of interest. Applying this to machine learning
(ML), and specifically supervised learning, for the task of protein
localisation prediction, these markers constitute the labelled training
data to use as input to a classification analyses. Defining well-known
residents, and obtaining labelled training data for ML analyses can be
time consuming, but it is important to define markers that are
representative of the multivariate data space and on which a classifier
will be trained and generated.
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} provides a
convenience function, \texttt{addMarkers}, to directly add markers to a
\texttt{MSnSet} object, as demonstrated in the code chunk below. These
marker sets can be accessed using the \texttt{pRolocmarkers()} function.
Marker sets are stored as a simple named vector in R, and originate from
in-house user-defined sets of markers or from previous published studies
(Gatto et al. 2014), which are continuosly updated and integrated.

@@

As demonstrated in (A. Christoforou et al. 2016) and illustrated in the
PCA plot above, the Golgi apparatus proteins display a dynamic pattern,
noting sets of Golgi marker proteins that are distributed amongst other
subcellular structures, an observation supported by microscopy. As such,
we are going to reset the annotation of Golgi markers to unknown using
the \texttt{fDataTounknown} function. It is often used to replace empty
strings (``'') or missing values in the markers definition to a common
definition of \emph{unknown localisation}.

@@

In general, the Gene Ontology (GO) (Ashburner et al. 2000), and in
particular the cellular compartment (CC) namespace are a good starting
point for protein annotation and marker definition. It is important to
note however that automatic retrieval of sub-cellular localisation
information, from
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} or
elsewhere, is only the beginning in defining a marker set for downstream
analyses. Expert curation is vital to check that any annotation added is
in the correct context for the the biological question under
investigation.

@@

With the aim of maximising the sub-cellular resolution and,
consequently, the reliability in protein sub-cellular assignments, we
follow the advice in (M. W. B. Trotter et al. 2010) and combine
replicated spatial proteomics experiments as described above. Indeed,
Trotter et al. have shown a significant improvement in
protein--organelle association upon direct combination of single
experiments, in particular when these resolve different subcellular
niches.

@@

The extraction of sub-cellular protein clusters can be difficult owing
to the limited number of marker proteins that exist in databases and
elsewhere. Furthermore, given the vast complexity of the cell, automatic
annotation retrieval does not always give a full representation of the
true sub-cellular diversity in the data. For downstream analyses, such
as supervised machine learning, it is desirable to obtain reliable
markers that cover as many sub-cellular niches as possible, as these
markers are directly used in the training phase of the ML
classification. We find that a lack of sub-cellular diversity in the
labelled training data leads to prediction errors, as unlabelled
instances can only be assigned to a class that exists in the training
data (L. M. Breckels et al. 2013). In such scenarios novelty detection
can be useful to identify data-specific sub-cellular groupings such as
organelles and protein complexes. The phenotype discovery (phenoDisco)
algorithm (L. M. Breckels et al. 2013) is one such method and is
available in
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}}. It is an
iterative semi-supervised learning method that combines the
classification of proteins on existing labelled data with the detection
of new clusters.

In addition to extracting new phenotypes, novelty detection methods are
also useful for confirming the presence of known or postulated clusters
in an unbiased fashion. For example, in (A. Christoforou et al. 2016)
the \texttt{phenoDisco} algorithm was used to confirm the data-specific
presence of the nucleus and nucleus sub-compartments. In the code chunk
below, we demonstrate how to do this analysis, highlighting some of the
optional arguments and parameters available for phenotype extraction and
give some advice on how to interpret the output.

As the \texttt{phenoDisco} algorithm is semi-supervised it uses both
labelled (markers) and unlabelled data to explore the data structure and
find new sub-cellular data clusters. Thus the first step is to define
some input labelled data i.e.~markers, that the algorithm will use as
input for the supervised learning aspect of the algorithm. As described
in (A. Christoforou et al. 2016) we define a set of markers to use as
input for the analyses that cover well-known residents from three
distinct organelle structures; the mitochondria, plasma membrane and ER,
and from three well-known and abundant protein complexes; the proteasome
and two ribosomal subunits, 40S and 60S. These input markers are stored
in the \texttt{featureData} column of \texttt{hl} where
\texttt{fcol = "phenoDisco.Input"}. We can use the convenience accessor
function \texttt{getMarkers} to print out a table of the markers
contained in this marker set. These initial markers were manually
curated using information from the UniProt database, the Gene Ontology
and the literature.

@@

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Thu Sep 29 13:42:12 2016 
## Added markers from  'pdres' marker vector. Thu Sep 29 13:42:13 2016 
##  MSnbase version: 1.99.2
\end{verbatim}

@@

Supervised machine learning, also known as classification, is an
essential tool for the assignment of proteins to distinct sub-cellular
niches. Using a set of labelled training examples i.e.~markers, we can
train a machine learning classifier to learn a mapping between the data
i.e.~the quantitative protein profiles, and a known localisation. The
trained classifier can then be used to predict the localisation of a
protein of unknown localisation, based on its observed protein profile.
To date, this method has been extensively used in spatial quantitative
proteomics to assign thousands of proteins to distinct sub-cellular
niches (A. Christoforou et al. 2016; A. J. Groen et al. 2014; M. W. B.
Trotter et al. 2010; Hall et al. 2009; Dunkley et al. 2006; Tan et al.
2009).

@@

\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4361037               0.3025581               0.3909670 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6803283               0.6104122               0.4318449 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4172007               0.5869100               0.9490415 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7940276               0.7083168               0.3151153 
##         Plasma membrane              Proteasome 
##               0.7224713               0.4169904
\end{verbatim}

@@

\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      86                     169                      88 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     297                     476                     103 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      24                     124                     523 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     229                     344                      39 
##         Plasma membrane              Proteasome                 unknown 
##                     315                     158                    2057
\end{verbatim}

@@

In addition to high quality MS-based quantitative proteomics data there
exist a number of other sources of information that are freely available
in the public domain that may be useful to assign a protein to its
sub-cellular niche. For example, imaging from immunofluorescence
microscopy, protein annotations and sequences, and protein-protein
interactions among others, represent a rich and vast source of
complementary information. We can integrate this auxiliary information
with our primary MS-based quantitative data using a paradigm known as
transfer learning (TL). The integration of data between different
technologies is one of the biggest challenges in computational biology
to date and the
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} package
provides functionality to do such analyses. We recently developed two
transfer learning algorithms using a \emph{k}-NN and SVM framework and
applied them to the task of protein localisation prediction (Breckels et
al. 2016). In this section we will begin with explaining the concept of
transfer learning and then show how to apply this in the frame of
spatial proteomics and protein localisation prediction.

@@

Now we have generated our auxiliary data we can use the \emph{k}-NN
implementation of transfer learning available in
\texttt{Biocpkg("pRoloc")} to integrate this with our primary MS-based
quantitative proteomics data using the functions
\texttt{knntlOptimisation} to estimate the free-parameters for the
integration, and \texttt{knntlClassification} to do the predictions. We
have shown that using transfer learning in the context of spatial
proteomics results in the assignment of proteins to sub-cellular niches
with a higher generalisation accuracy than using standard supervised
machine learning with a single source of information (Breckels et al.
2016).

The first step, as with any machine learning algorithm, is to optimise
any free paramaters of the classifier. For the \emph{k}-NN TL classifier
there are two sets of parameters that need optimising: the first set are
the $k$'s for the primary and auxiliary data sources required for the
nearest neighbour calculations for each data source. The second set of
parameters (which is noted by a vector of $\theta$ weights) that require
optimising are the class weights, one per subcellular niche, that
control the proportion of primary and auxiliary data to use for
learning. A weight can take any real value number between 0 and 1. A
weight of $\theta = 1$ indicates that all weight is given to the primary
data (and this implicitly implies that a weight of $1 - \theta$ is given
to the auxiliary data, similarly a weight of $\theta = 0$ implies that
all weight is given to the auxiliary data (so 0 is given to the primary
source). If we conduct a parameter seach and test weights
$\theta = {0, 1/3, 2/3, 1}$ for each class, and if we have, for example
10 subcellular niches, this will result in \texttt{4\^{}10} different
combinations of parameters to test. The parameter optimisation is
therefore time consuming and as such we recommend you make use of a
computing cluster (code and submissing scripts are also available in the
supporting information). The markers in the \texttt{hl} dataset contain
14 subcellular classes. If we examine these markers and classes on the
PCA plot above we can see that in particular the two ribosomes and two
nuclear compartments are highly separated along the first two
components, this is also evident from the profiles plot which gives us a
good indication that these subcellular niches are well-resolved in the
hyperLOPIT dataset. Transfer learning is particularly useful for classes
that are not as well separated, we find that subcellular niches that are
well-separated under hyperLOPIT and LOPIT obtain a class score of 1
(i.e.~use only primary data from transfer learning (Breckels et al.
2016)). Therefore, for the optimisation stage of the analyses we can
already infer a subcellular class weight of 1 for these niches and only
optimise over the remaining organelles. This can signifciantly cut down
optimisation time as by removing these 4 classes from the optimisation
(and not the classification) we only have \texttt{4\^{}10} class weight
combinations to consider instead of \texttt{4\^{}14} combinations.
